The first predicates \SAMEDAILYSLOT, , same slot enforce common restrictions on the start times of the targeted sessions (e.g.,
sessions starting the same day). Additionally, any start time interval may
be forbidden by passing its start and end points as parameters to predicate
forbidden period. Predicates at most daily and at most weekly upper-bound the number of sessions scheduled daily or weekly within the given
time interval. sequenced is a n-ary predicate (n >= 2) which constrains the
latest session of the i-th e-map to end before the earliest session of i + 1-th e-map (i = 1..n - 1). Predicate weekly ensures sessions are scheduled
weekly without presuming any particular sequencing. Predicate no overlap
ensures sessions do not overlap in time and is typically used to model disjunctive resources. Predicate travel factors in any travel time incurred between
consecutive sessions hosted in distant rooms. The travel time matrix is a parameter of the predicate. same rooms, same students and same lecturers
require that sessions be assigned to the same set of rooms, students or lecturers. Predicate adjacent rooms require that sessions be hosted in adjacent
rooms based on an adjacency graph passed as a parameter. Lastly, predicate
lecturer distribution distributes the volumes of sessions represented by
the different e-map arguments among different lecturers. Lecturers and session volumes are parameters of the predicate.


\textbf{\ADJACENTROOMS}((e,S'),K$_1$,$\dots$,K$_n$) with $ K_i \subseteq \ROOM, i : 1..n $ where K\_i