\coco{OLD}
The first predicates 
\texttt{\SAMEDAILYSLOT},
\ldots,
%\texttt{\SAMEWEEKDAY},
%\texttt{\SAMEWEEKLYSLOT},
%\texttt{\SAMEWEEK},
%\texttt{\SAMEDAY} and
\texttt{\SAMESLOT}
enforce common restrictions on the start times of the targeted sessions (e.g., sessions starting the same day).
Additionally,
any start time interval may be forbidden 
by passing its start and end points 
as parameters to 
predicate \texttt{\FORBIDDENPERIOD}.
Predicates \texttt{\ATMOSTDAILY}
and
\texttt{\ATMOSTWEEKLY}
upper-bound
the number of sessions
scheduled daily or weekly
within the given time interval.
\texttt{\SEQUENCED}
is a n-ary predicate ($n\geq2$)
which constrains
the latest session of the $i$-th e-map 
to end before
the earliest session of $i+1$-th e-map ($i=1..n-1$).
Predicate 
\texttt{\WEEKLY}
ensures sessions
are scheduled weekly
without presuming any particular sequencing.
Predicate
\texttt{\NOOVERLAP}
ensures sessions do not overlap in time
and is typically used to model disjunctive resources.
Predicate \texttt{\TRAVEL}
factors in any travel time
incurred between consecutive sessions
hosted in distant rooms.
The travel time matrix is a parameter of the predicate.
\texttt{\SAMEROOMS},
\texttt{\SAMESTUDENTS}
and
\texttt{\SAMETEACHERS}
require that sessions be assigned to the same set of rooms,
students or lecturers.
Predicate 
\texttt{\ADJACENTROOMS}
require that sessions be hosted in 
adjacent rooms 
based on an adjacency graph passed as a parameter.
Lastly, 
predicate \texttt{\TEACHERDISTRIBUTION}
distributes the volumes of sessions represented by the different e-map arguments 
among different lecturers. Lecturers and session volumes are parameters of the predicate.

